---
title: "p8105_hw2_gem2176"
author: "Giuliana Motta"
date: "2026-10-05"
output: github_document
---

Libraries
```{r}
library(tidyverse)
library(dplyr)
library(readxl)
```


# Problem 1

#### Reading and cleaning NYC transit data
```{r}
nyc_transit_df =
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |> 
  janitor :: clean_names() |> 
  select(line, station_name, station_latitude, station_longitude,
         route1:route11, entry, vending, entrance_type, ada) |>
  mutate(
    entry = 
      case_match( 
        entry,
        "YES" ~ TRUE,
        "NO"~ FALSE),
    entry = as.logical(entry))
```
This data set describes the NYC subway station entrances and exits. For each entrance and exit, the data set describes the line, the station, the latitude and longitude of the station, the routes that pass through the station (up to 11 routes), whether entry is allows, whether the entry is with or without vending, the type of entrance, and whether the entrance is ADA compliant. Thus far, the data has been cleaned to only include variables of interest and to transform the entry variable from a character variable to a vector variable. The current data frame has 19 columns and 1868 rows. The data is not tidy because the route number and name are in wide format instead of long format. 

#### Calculating the number of distinct stations
```{r}
nrow(distinct(nyc_transit_df, line, station_name))
```
There are 465 distinct stations in NYC


#### Assessing the number of stations that are ADA compliant
```{r}
nyc_transit_df |> 
  filter(ada == TRUE) |> 
  distinct(line, station_name) |>
  nrow()
```
There are 84 stations that are ADA compliant 


#### Calculating the proportion of station entrances and exits without vending that allow entrance
```{r}
denominator <- nyc_transit_df |> 
  filter(vending == "NO") |> 
  nrow()

numerator <- nyc_transit_df |> 
  filter(vending == "NO") |> 
  filter(entry == TRUE) |> 
  nrow()

100*(numerator/denominator)
```
37.71% of station entrances and exits without vending allow entrance. 


#### Reformatting data so that route number and route name are distinct variables. 
```{r}
nyc_transit_tidy_df = 
  nyc_transit_df |> 
  mutate(
    route8 = as.character(route8), 
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)) |> 
  pivot_longer(
    route1:route11, #columns 
    names_to = "route_number", 
    names_prefix = "route",
    values_to = "route_name")

#Stations that serve the a train
nyc_transit_tidy_df |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |>
  nrow()

#A train stations that are ADA compliant
nyc_transit_tidy_df |> 
  filter(route_name == "A") |>
  filter(ada == TRUE) |> 
  distinct(line, station_name) |>
  nrow()
```
There are 60 stations that serve the a train and 17 of these stations are ada compliant.

# Problem 2

#### Reading in Mr. Trash Wheel Dataset
```{r}
mr_trash_df = 
  read_excel("data/202610 Trash Wheel Collection Data.xlsx", 
                      sheet = "Mr. Trash Wheel",
                      range = "A2:N709", 
                      col_names = TRUE,
                      na = "") |> 
  janitor :: clean_names() |> 
  mutate(
    sports_balls = as.integer(sports_balls),
    trash_wheel = "mr_trash",
    year = as.numeric(year))
```

#### Reading in Prof. Trash Wheel Dataset
```{r}
prof_trash_df = 
  read_excel("data/202610 Trash Wheel Collection Data.xlsx", 
                      sheet = "Professor Trash Wheel",
                      range = "A2:M134", 
                      col_names = TRUE,
                      na = "") |> 
  janitor :: clean_names() |> 
  mutate(trash_wheel= "prof_trash",
         year = as.numeric(year))

```

#### Reading in Prof. Trash Wheel Dataset 
```{r}
gwynn_trash_df = 
  read_excel("data/202610 Trash Wheel Collection Data.xlsx", 
                      sheet = "Gwynnda the Good Wheel of the W",
                      range = "A2:L351", 
                      col_names = TRUE,
                      na = "") |> 
  janitor :: clean_names() |> 
  mutate(trash_wheel= "gwynn_trash", 
         year= as.numeric(year))
```


#### Produce a single tidy dataset
```{r}
trash_tidy_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwynn_trash_df) |> 
  pivot_longer(
    plastic_bottles:sports_balls, #columns
    names_to = "litter_type", 
    values_to = "litter_count")
  )
```

#### Calculating the total weight of trash collected by Professor Trash Wheel and the total number of cigarette butts collected by Gwynnda in June of 2022. 
```{r}
ptw_sum = sum(mr_trash_df$weight_tons)

gtw = 
  gwynn_trash_df |> 
  filter(year == 2022) |> 
  filter(month == "June")

gtw_sum = sum(gtw$cigarette_butts)
```


This data set describes the items removed by three trash wheels (Mr. Trash, Prof. Trash, and Gwynn Trash) that remove litter from Baltimore's inner harbor. Observations are recorded daily over `r nrow(trash_tidy_df)` days. Variables include weight (tons) and volume (cubic yards) of the litter picked, the trash wheel that picked the litter, and counts for various forms of litter including plastic bottles, polystyrene and cigarette butts. The number of homes powered from this effort is also recorded. For the available data the total weight of trash collected by Professor Trash Wheel was `r ptw_sum` tons and the total number of cigarette butts collected by Gwynnda in June of 2022 was `r format(gtw_sum, scientific= FALSE)`.


# Problem 3

#### Importing, cleaning, and tidying the dataset of baseline demographics
```{r}
baseline_df =
  read_csv("data/data_mci/MCI_baseline.csv",
           skip = 1,
           na = c("NA", "", ".")) |> 
  janitor :: clean_names() |> 
  mutate(
    apoe4 = 
      case_match( 
        apoe4,
         1 ~ TRUE,
         0 ~ FALSE),
    apoe4 = as.logical(apoe4), 
    sex = 
      case_match( 
        sex,
         1 ~ "male",
         0 ~ "female"))


baseline_omit_df =
  read_csv("data/data_mci/MCI_baseline.csv",
           skip = 1,
           na = c("NA", "", ".")) |> 
  janitor :: clean_names() |> 
  mutate(
    apoe4 = 
      case_match( 
        apoe4,
         1 ~ TRUE,
         0 ~ FALSE),
    apoe4 = as.logical(apoe4), 
    sex = 
      case_match( 
        sex,
         1 ~ "male",
         0 ~ "female")) |>
  filter(is.na(age_at_onset)|current_age<age_at_onset)
  
  
```

#### Calculating the number of participants who developed MCI, the average baseline age, and the proportion of women in the study who are APOE4 carriers. 
```{r}

mci_develop = 
  baseline_omit_df |> 
  filter(!is.na(age_at_onset))

mean(baseline_omit_df$current_age)

wom =
  baseline_omit_df |> 
  filter(sex=="female") 

wom_apoe_4 =
  baseline_omit_df |> 
  filter(sex=="female") |> 
  filter(apoe4== TRUE)

nrow(wom_apoe_4)/nrow(wom)
```


This data set represents baseline demographic data at study baseline. Variables include: current age, sex, education, whether the participant is an APOE4 carrier, and age of mild cognitive impairment (MCI) onset. Data was cleaned, characterizing "." as NA values, and designating APOE4 and sex a logical and character form, respectively. `r nrow(baseline_df)` participants were recruited. `r nrow(baseline_df)-nrow(baseline_omit_df)` participants who already had had mild cognitive impairment at baseline were excluded from the analysis, leading to a total of `r nrow(baseline_omit_df)` participants. Of these participants, `r nrow(mci_develop)` developed MCI. The average baseline age was `r round(mean(baseline_omit_df$current_age),2)` years and the proportion of women in the study who are APOE4 carriers is `r nrow(wom_apoe_4)/nrow(wom)`. 

#### Importing, cleaning, and tidying the dataset of longitudinally observed biomarker values

```{r}
mci_amyloid_df_long =
  read_csv("data/data_mci/mci_amyloid.csv",
           skip = 1,
           na = c("NA", "", "Na")) |> 
  janitor :: clean_names() |> 
  rename(id = study_id ) |> 
  pivot_longer(
    baseline:time_8, #columns
    names_to = "time", 
    values_to = "amyloid ratio")

mci_amyloid_df_wide =
  read_csv("data/data_mci/mci_amyloid.csv",
           skip = 1,
           na = c("NA", "", "Na")) |> 
  janitor :: clean_names() |> 
  rename(id = study_id )
```
Data was imported from a CSV file and cleaned, ensuring that values inserted as "Na" and "NA" were coded as NA. The data was tidied by converting from wide format to long format. 

#### Checking whether some participants appear only in the

```{r}
#Performing a full join to identify the ids that do not match 

baseline_missing=
  anti_join(baseline_omit_df,mci_amyloid_df_wide)

amyloid_missing=
  anti_join(mci_amyloid_df_wide,baseline_omit_df)

```
`r nrow(baseline_missing)` participant ids appear only in the amyloid dataset and `r nrow(amyloid_missing)` participant ids appear only in the baseline dataset. These participants were excluded from the dataset. 

#### Removing participants that appear only in one dataset 
```{r}
combined_mci_wide_df=
  inner_join(baseline_omit_df, mci_amyloid_df_wide)

combined_mci_long_df=
  inner_join(baseline_omit_df, mci_amyloid_df_long)
```
This dataset combines the demographic and biomarker datasets so that only participants who appear in both datasets are retained. The resulting data has `r nrow(combined_mci_wide_df)` participants.

#### Exporting data
```{r}
write_csv(combined_mci_long_df, file = "mci_combined.csv")
```

